---
title: "Descriptive & Inferential Analysis of a Jungian Sandplay VR Project"
subtitle: "Normality testing of patient treatment - Report"
author: "Gareth Burger &amp; James Bunt"
date: "`r format(Sys.time(), '%a, %d %B %y at %X')`"
fontsize: "10pt"
output: 
  html_document:
    toc: true
    toc_title: "Table Of Contents"
    toc_float: true
    toc_depth: 3
    fig_caption: false
    lof: true
    number_sections: true
    code_folding: hide
    theme: paper
---

```{r setup and library load, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
cat('\014')
```

**WIP**\
*We can rename this README or make the main R Notebook file a completely new file if needed*\
*2-2.5 pages of text needed, not more than 1200 words excluding plots and tables*\

# Abstract
*NB to be succinct and clear here*

## Aim and rationale

## Participants and setting

## Experiment design

## Results gathering

## Major findings

## Findings/implications


# Introduction

## Topic and context

## Rationale

*2. Formulate a single hypothesis test to be used to compare the effectiveness of the approaches used during the experiment.*\

## Hypothesis


# Method

## Participants

## Design

## Materials

## Procedure

### Load Data
```{r Load Data, echo=TRUE}
# load data from csv
file_name = "./data/daie_ca3_data_13.csv"
study_data <- read.csv(file_name)
```

*1. Determine whether the data provided are appropriate for the test(s) available and whether any data cleaning is required.*\

### Inspect Data Structure
```{r Inspect Head, Tail & Structure, echo=TRUE}
# show data structure
str(study_data)

# show first N
head(study_data, 10)

# show last N
tail(study_data, 10)

# show a paged table containing all the data
rmarkdown::paged_table(study_data)
```

### Clean Data
```{r}
# install.packages("tidyverse")
# install.packages("dplyr")

# load and attach tidyverse and dplyr packages
library(tidyverse)
library(dplyr)

View(study_data)

# variable types
glimpse(study_data)
class(study_data$post_trial_cpss)

# find distinct data values for columns
unique(study_data$gender) # incorrect spelling of value "mal", should be "Male"
unique(study_data$test_group) # value of "" is present, means missing data

# select variables (selects all valid data - see count is 147 row i.e. 3 rows have invalid/missing data)
study_data %>%
  select(gender, test_group, ends_with("cpss")) %>%
  filter(gender %in% c("Male", "Female") 
         & test_group %in% c("Static", "Control", "Animated") 
         & post_trial_cpss > 0, post_trial_cpss < 10)

# identify missing data
#mean(study_data$pre_trial_cpss, na.rm = TRUE) # remove not available observations from pre_trial_cpss variable
mean(study_data$pre_trial_cpss) #6.0618
mean(study_data$post_trial_cpss) #5.4154
mean(study_data$pre_trial_or) #6.072133
mean(study_data$post_trial_or) #5.3748
# all means were able to be calculated, thus there's no missing data in those columns

# filter observations of variables that have missing values (does not identify the 1 line with missing data, unsure why yet)
study_data %>%
  select(gender, test_group, pre_trial_cpss, post_trial_cpss, pre_trial_or, post_trial_or) %>%
  filter(!complete.cases(.)) # complete.cases not picking up row 42 '' test_group

# check for duplicates (no dupes)
study_data %>% distinct() # 150 rows returned, so no duplicates
duplicated(study_data) # all logical vector values are FALSE i.e. no duplicates

# continue here...
```

### Mean and Standard Deviation
```{r Mean & Median, echo=TRUE}
pre_trial_cpss_mean <- mean(study_data$pre_trial_cpss)
pre_trial_cpss_std <- sd(study_data$pre_trial_cpss)

post_trial_cpss_mean <- mean(study_data$post_trial_cpss)
post_trial_cpss_std <- sd(study_data$post_trial_cpss)

pre_trial_or_mean <- mean(study_data$pre_trial_or)
pre_trial_or_std <- sd(study_data$pre_trial_or)

post_trial_or_mean <- mean(study_data$post_trial_or)
post_trial_or_std <- sd(study_data$post_trial_or)
```

pre_trial_cpss: M = `r round(pre_trial_cpss_mean, 2)`, SD = `r round(pre_trial_cpss_std, 2)`

post_trial_cpss: M = `r round(post_trial_cpss_mean, 2)`, SD = `r round(post_trial_cpss_std, 2)`

pre_trial_or: M = `r round(pre_trial_or_mean, 2)`, SD = `r round(pre_trial_or_std, 2)`

post_trial_or: M = `r round(post_trial_or_mean, 2)`, SD = `r round(post_trial_or_std, 2)`

### Data Summary
A summary of the data is shown below:

```{r Summary, echo=TRUE}
summary(study_data)
```

*3. Determine if the data meet the assumptions required by any statistical test.*\

### Three Tests for Normality
[How to Test for Normality in R](https://www.statology.org/test-for-normality-in-r/)

#### Method 1: Using Histogram / Density Plot
```{r Histogram & Density Plot - pre_trial_cpss, echo=TRUE}
# draw histogram for pre_trial_cpss
hist(study_data$pre_trial_cpss, main = "pre_trial_cpss", xlab = "Measured Score", ylab = "Frequency", breaks = 6)
```

```{r Histogram & Density Plot - post_trial_cpss, echo=TRUE}
# draw histogram for post_trial_cpss
hist(study_data$post_trial_cpss, main = "post_trial_cpss", xlab = "Measured Score", ylab = "Frequency", breaks = 6)
```

```{r Histogram & Density Plot - pre_trial_or, echo=TRUE}
# draw histogram for pre_trial_or
hist(study_data$pre_trial_or, main = "pre_trial_or", xlab = "Measured Score", ylab = "Frequency", breaks = 6)
```

```{r Histogram & Density Plot - post_trial_or, echo=TRUE}
# draw histogram for post_trial_or
hist(study_data$post_trial_or, main = "post_trial_or", xlab = "Measured Score", ylab = "Frequency", breaks = 7)
```

#### Method 2: Using QQ Plot (Version 1 - using in-built qqnorm())
*visual inspection test*\
*samples are close to the qq line so gives a good indication that the data is normally distributed*\
*though they don't all fall exactly on the line so we'll do the next test to further investigate*\
```{r QQ Plot - pre_trial_cpss, echo=TRUE}
# perform qqnorm test on pre_trial_cpss
qqnorm(study_data$pre_trial_cpss, pch = 2, frame = FALSE)
qqline(study_data$pre_trial_cpss, col = "darkorchid", lwd = 1)
```

```{r QQ Plot - post_trial_cpss, echo=TRUE}
# perform qqnorm test on post_trial_cpss
qqnorm(study_data$post_trial_cpss, pch = 2, frame = FALSE)
qqline(study_data$post_trial_cpss, col = "darkorchid", lwd = 1)
```

```{r QQ Plot - pre_trial_or, echo=TRUE}
# perform qqnorm test on pre_trial_or
qqnorm(study_data$pre_trial_or, pch = 2, frame = FALSE)
qqline(study_data$pre_trial_or, col = "darkorchid", lwd = 1)
```

```{r QQ Plot - post_trial_or, echo=TRUE}
# perform qqnorm test on post_trial_or
qqnorm(study_data$post_trial_or, pch = 2, frame = FALSE)
qqline(study_data$post_trial_or, col = "darkorchid", lwd = 1)
```

#### Method 3: Statistical Test: Shapiro-Wilk
*the threshold for normality is 0.05*\
*if the p-value is above the threshold then we can accept the data to be normal*\
*if the p-value is close to the boundary, we want to test via a fourth method*\
```{r Shapiro-Wilk - pre_trial_cpss, echo=TRUE}
# perform shapiro-wilk test on pre_trial_cpss
pre_trial_cpss_sw <- shapiro.test(study_data$pre_trial_cpss)
pre_trial_cpss_sw
```
A `r pre_trial_cpss_sw$method` was conducted on the pre_trial_cpss data.\
From the output obtained we can assume normality as the p-value (p=`r round(pre_trial_cpss_sw$p.value, 4)`) is greater than 0.05.

```{r Shapiro-Wilk - post_trial_cpss, echo=TRUE}
# perform shapiro-wilk test on post_trial_cpss
post_trial_cpss_sw <- shapiro.test(study_data$post_trial_cpss)
post_trial_cpss_sw
```
A `r post_trial_cpss_sw$method` was conducted on the post_trial_cpss data.\
*data still needs to be cleaned*\
From the output obtained we can not assume normality as the p-value (p=`r format.pval(post_trial_cpss_sw$p.value, 5, eps = 0.001)`) is less than 0.05.

```{r Shapiro-Wilk - pre_trial_or, echo=TRUE}
# perform shapiro-wilk test on pre_trial_or
pre_trial_or_sw <- shapiro.test(study_data$pre_trial_or)
pre_trial_or_sw
```
A `r pre_trial_or_sw$method` was conducted on the pre_trial_or data.\
From the output obtained we can assume normality as the p-value (p=`r round(pre_trial_or_sw$p.value, 4)`) is greater than 0.05.

```{r Shapiro-Wilk - post_trial_or, echo=TRUE}
# perform shapiro-wilk test on post_trial_or
post_trial_or_sw <- shapiro.test(study_data$post_trial_or)
post_trial_or_sw
```
A `r post_trial_or_sw$method` was conducted on the post_trial_or data.\
From the output obtained we can assume normality as the p-value (p=`r round(post_trial_or_sw$p.value, 4)`) is greater than 0.05.


# Results

*4. Provide descriptive statistics (graphs and tables) for any assumptions made.*\

## Descriptive statistics
*plots - box, bar, scatter, qq plots*

## Inferential statistics
*the inference that we are doing on our data requires that the data is normally distributed*\

*6. Determine the 95% confidence interval for the population mean of each group, and the 95% confidence interval for the difference between the means of any two groups for a variable of your choice.*\

95% confidence interval = point estimate +- 1.96 * standard error

SE = sqrt( p(1-p)/n )  - slides 19-20 of chapter 5

p = point estimate +/- 1.96 * SE
  = PE +/- ***
We are 95% confident that [PE + x, PE - x] of all animated group


*7. Determine the degree of correlation between any explanatory and response variable of your choice.*\

*do student t test - see diamond quality video*

## Statistical tests
[7 Types of Statistical Analysis: Definition and Explanation](https://www.analyticssteps.com/blogs/7-types-statistical-analysis-definition-explanation)

## Magnitude and direction of results


# Discussion

*5. Analyse the data to provide the hypothesis testing conclusion.*\

## Outline findings and relation to the hypothesis

## Limitations (if confounding variables are clearly identified by your group)


# References
- *list here*
-
-