---
title: "Descriptive & Inferential Analysis of a Jungian Sandplay VR Project"
subtitle: "Normality testing of patient treatment"
author: "Gareth Burger &amp; James Bunt"
date: "`r format(Sys.time(), '%a, %d %B %y at %X')`"
fontsize: "10pt"
output: 
  html_document:
    toc: true
    toc_title: "Table Of Contents"
    toc_float: true
    toc_depth: 3
    fig_caption: false
    lof: true
    number_sections: true
    code_folding: hide
    theme: paper
---

**WIP**

*We can rename this README or make the main R Notebook file a completely new file if needed*

*2-2.5 pages of text needed, not more than 1200 words excluding plots and tables*

```{r setup and library load, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
cat('\014')
```


# Abstract
*NB to be succinct and clear here*

## Aim and rationale



## Participants and setting



## Experiment design



## Results gathering

### Load Data
```{r Load Data, echo=TRUE}
# load data from csv
file_name = "./data/daie_ca3_data_13.csv"
study_data <- read.csv(file_name)
```

### Inspect Data Structure
```{r Inspect Head, Tail & Structure, echo=TRUE}
# show data structure
str(study_data)

# show first N
head(study_data, 10)

# show last N
tail(study_data, 10)

# show a paged table with all the data
rmarkdown::paged_table(study_data)
```

### Mean and Standard Deviation
```{r Mean & Median, echo=TRUE}
pre_trial_cpss_mean <- mean(study_data$pre_trial_cpss)
pre_trial_cpss_std <- sd(study_data$pre_trial_cpss)

post_trial_cpss_mean <- mean(study_data$post_trial_cpss)
post_trial_cpss_std <- sd(study_data$post_trial_cpss)

pre_trial_or_mean <- mean(study_data$pre_trial_or)
pre_trial_or_std <- sd(study_data$pre_trial_or)

post_trial_or_mean <- mean(study_data$post_trial_or)
post_trial_or_std <- sd(study_data$post_trial_or)
```

pre_trial_cpss: M = `r round(pre_trial_cpss_mean, 2)`, SD = `r round(pre_trial_cpss_std, 2)`

post_trial_cpss: M = `r round(post_trial_cpss_mean, 2)`, SD = `r round(post_trial_cpss_std, 2)`

pre_trial_or: M = `r round(pre_trial_or_mean, 2)`, SD = `r round(pre_trial_or_std, 2)`

post_trial_or: M = `r round(post_trial_or_mean, 2)`, SD = `r round(post_trial_or_std, 2)`

### Data Summary
A summary of the data is shown below:

```{r Summary, echo=TRUE}
summary(study_data)
#boxplot(student_data$height_cm~student_data$gender)
```



## Major findings



## Findings/implications



# Introduction

## Topic and context



## Rationale



## Hypothesis



# Method

## Participants



## Design



## Materials



## Procedure



# Results

## Descriptive statistics



## Inferential statistics



## Statistical tests



## Magnitude and direction of results



# Discussion

## Outline findings and relation to the hypothesis



## Limitations (if confounding variables are clearly identified by your group)



# References
- *list here*
-
-