---
title: "Descriptive & Inferential Analysis of a Jungian Sandplay VR Project"
author: "Gavin Burger & James Bunt"
date: "`r format(Sys.time(), '%a, %d %B %y at %X')`"
fontsize: "10pt"
output: 
  html_document:
    toc: true
    toc_title: "Table Of Contents"
    toc_float: true
    toc_depth: 2
    fig_caption: false
    lof: true
    number_sections: true
    code_folding: hide
    theme: paper
---

```{r setup and library load, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
cat('\014')
```

#Load & Inspect Data

## Load Data

```{r Load Data, echo=TRUE}
file_name = "./data/daie_ca3_data_13.csv"
study_data <- read.csv(file_name)
```

# install.packages("tidyverse")
library(tidyverse)
view(study_data)

# variable types
glimpse(study_data)
unique(study_data)
unique(study_data$gender)
unique(study_data$test_group)

# filter variables
study_data %>%
  select(gender, test_group, ends_with("cpss")) %>%
  filter(gender %in% c("Male, "Female") & test_group %in% c("Static", "Animated", "Control") & post_trial_cpss < 10) %>%

# change variable to factor variable 
study_data$gender <- as.factor(study_data$gender)
study_data$test_group <- as.factor(study_data$test_group)

# Missing data


## Inspect Data Structure
```{r Inspect Head, Tail & Structure, echo=TRUE}
# show structure
str(study_data)
# show first N
head(study_data, 10)
# show last N
tail(study_data, 5)
# show a "prettier" paged table
rmarkdown::paged_table(study_data)
# show a "prettier" paged table but ONLY first 8 rows
head_study_data <- head(study_data, 8)
rmarkdown::paged_table(head_study_data)
# show a "prettier" paged table using kable() with caption and decimal digit max length and alignment = centre
knitr::kable(study_data,
  align = c('c'), # a vector with 'c', 'l', or 'r' values
  digits = 2,
  caption = "Source data table")
```

             
## Mean, Median & Summary

```{r Mean & Median, echo=TRUE}
shoe_size_mean <- mean(study_data$show_size_eu)
shoe_size_std <- sd(student_data$show_size_eu)
age <- 40
```

The value of age is `r age`

We can access variables from inside code chunks by wrapping the variable in a *mini-chunk* of R as shown on the next line.

Median shoe size statistics were generated (M =`r round(shoe_size_mean, 2)`, SD = `r round(shoe_size_std, 2)`). 

It can be useful to create a summary of the data for quick analysis. A summary of the data is shown below:

```{r Summary, echo=TRUE}
summary(student_data)
#boxplot(student_data$height_cm~student_data$gender)
```



